\pdfoutput=1

\documentclass[aps,prd,reprint,twocolumn,showpacs,nofootinbib,superscriptaddress,floatfix]{revtex4-1}

\usepackage{amsmath,amssymb,amstext}
\usepackage[usenames,svgnames]{xcolor}
\usepackage{hyperref,graphicx}
\usepackage{subfigure}

\hypersetup{
    colorlinks=true,
    urlcolor=SteelBlue,
    linkcolor=red,
    citecolor=blue,
}

\renewcommand{\vec}[1]{\mathbf{#1}}

\begin{document}
\title{Power spectrum and particle production\\ in galileon bouncing cosmologies}

\author{David A. Dobre}%
\email{ddobre@sfu.ca}
\affiliation{Department of Physics, Simon Fraser University,\\
8888 University Drive, Burnaby, British Columbia V5A 1S6, Canada}

\author{Andrei V. Frolov}%
\email{frolov@sfu.ca}
\affiliation{Department of Physics, Simon Fraser University,\\
8888 University Drive, Burnaby, British Columbia V5A 1S6, Canada}

\author{Jos\'{e} T. G\'{a}lvez Ghersi}%
\email{joseg@sfu.ca}
\affiliation{Department of Physics, Simon Fraser University,\\
8888 University Drive, Burnaby, British Columbia V5A 1S6, Canada}
\affiliation{Perimeter Institute for Theoretical Physics,\\
31 Caroline Street North, Waterloo, Ontario, N2L 2Y5, Canada}

\author{Alexander Vikman}%
\email{vikman@fzu.cz}
\affiliation{Institute of Physics, the Academy of Sciences of the Czech Republic,\\
Na Slovance 2, 182 21 Prague 8, Czech Republic}
\date{\today}


\begin{abstract}
In this paper, we study the main features of the scalar and tensor power spectrum of primordial fluctuations during the classically stable bouncing phase recently proposed by Ijjas and Steinhardt in Ref.~\cite{Ijjas:2016tpn}. Following the suggested background history, we evaluated the tensor and scalar spectra during the bouncing phase characterized by the violation of null (and strong) energy condition. We found that the change of the speed of sound is the cause of the dominance of the tensor power spectrum over the scalar part through most of the bounce. Moreover, we observe that none of the spectra evaluated across the bouncing phase is scale invariant. In addition to this, we present our results for particle production by showing the evolution of the occupation number of scalar fluctuations through the bounce.    
\end{abstract}

\maketitle


\section{Introduction}
Bouncing cosmologies are typically studied as effective field manifestations of many high energy theories of grand unification with extended symmetry, as shown in Refs.~\cite{Ashtekar:2006rx, Poplawski:2011jz}. Moreover, the production of healthy bouncing trajectories is an important accomplishment in order to produce a sensible expansion history free of singularities. A satisfactory non-singular bounce might not require from a complete description of the theory in the ultraviolet limit -- including Quantum Gravity -- while remaining coherent with the observed stages of homogeneity and isotropy seen in Cosmic Microwave Background (CMB). It is possible to find recent progress in the approach described in Refs.~\cite{Ijjas:2016tpn, Ijjas:2016vtq}, wherein the expansion history and the running of the kinetic couplings are used as inputs to describe the full dynamics of a cubic galileon. A valuable feature of this approach is that it is exactly the inverse when compared to other methods in which a field with a non-canonical kinetic term is used as a source (for example, in Refs.~\cite{Deffayet:2010qz, Easson:2011zy}) of the bouncing expansion history.  

The study of scalar and tensor perturbations is remarked as important across the existing literature: it adds additional constraints to the notions of a stable bouncing trajectory by avoiding all forms of instabilities such as $c_s^2<0$, tachyonic and gradient. Nevertheless, the dynamical properties of these fluctuations are not fully explored. This might be due to the cumbersome time dependence of the second-order expanded action, or because of the need of expensive computational power to resolve at higher frequencies. In this paper, we compute the numerical evolution of the scalar and tensor modes by implementing a simplified version of the separation technique described in Ref.~\cite{Ghersi:2016gee}, where each of the equations of motion of each mode is separated in two parts: one equation -- with a reduced oscillation frequency -- evolves the amplitude and another first-order equation evaluates the phase. After this separation, the numerical treatment of the system not only permits to visualize the power spectrum of curvature scalar and tensor fluctuations at a given instant of time, but all across the bouncing phase with sufficient resolution in Fourier domain. Furthermore, the dynamical evaluation of the spectra throughout the bounce is relevant since the (de-)accelerating background has a continuous effect on the perturbations. A visible effect of this acceleration is the dynamical production of particles, which can also be studied as the background evolves and will also be reflected in the particular features of the spectra.              

The plan of this paper is as follows, in section \ref{secII}, we review the features of the bouncing model explored in Ref.~\cite{Ijjas:2016tpn} and the suggested bouncing trajectory. We will study the forbidden regions of phase space for the cubic galileon, where the scalar perturbations are unstable. In addition to this, we will also find the location of the bouncing trajectories in phase space. In section \ref{secIII}, we describe the the action, equations of motion and initial conditions required to find the spectrum of scalar and tensor fluctuations. It is widely known that the choice of initial conditions is not unique, our choice was a state of instantaneous minimal energy previous to the bouncing stage in order to have properly defined initial conditions at all wavelengths. We also review the standard definition of occupancy number as stated in Ref.~\cite{Barnaby:2010ke}. In section \ref{secIV}, we show our results for the power spectra of primordial scalar and tensor fluctuations. We present two major results: (I) the power spectra are not scale invariant and (II) the scalar spectrum only dominates over the tensor by the end of the bouncing phase. In this section, we exhibit our results for the scalar particle creation and its effects in the spectrum. Finally, we discuss and conclude.      

\section{Background}\label{secII}
\section{Scalar and tensor fluctuations}\label{secIII}
\section{Power spectrum of scalar and tensor fluctuations}\label{secIV}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{99}

\bibitem{Ijjas:2016tpn} 
  A.~Ijjas and P.~J.~Steinhardt,
  \textit{``Classically stable nonsingular cosmological bounces,''}
  Phys.\ Rev.\ Lett.\  {\bf 117}, no. 12, 121304 (2016)
  doi:10.1103/PhysRevLett.117.121304

\bibitem{Ashtekar:2006rx} 
  A.~Ashtekar, T.~Pawlowski and P.~Singh,
  \textit{``Quantum nature of the big bang,''}
  Phys.\ Rev.\ Lett.\  {\bf 96}, 141301 (2006)
  doi:10.1103/PhysRevLett.96.141301
  [gr-qc/0602086].
  
\bibitem{Poplawski:2011jz} 
  N.~J.~Poplawski,
  \textit{``Nonsingular, big-bounce cosmology from spinor-torsion coupling,''}
  Phys.\ Rev.\ D {\bf 85}, 107502 (2012)
  doi:10.1103/PhysRevD.85.107502
  [arXiv:1111.4595 [gr-qc]].
  
\bibitem{Ijjas:2016vtq} 
  A.~Ijjas and P.~J.~Steinhardt,
  \textit{``Fully stable cosmological solutions with a non-singular classical bounce,''}
  Phys.\ Lett.\ B {\bf 764}, 289 (2017)
  doi:10.1016/j.physletb.2016.11.047
  [arXiv:1609.01253 [gr-qc]].
  
  \bibitem{Deffayet:2010qz} 
  C.~Deffayet, O.~Pujolas, I.~Sawicki and A.~Vikman,
  \textit{``Imperfect Dark Energy from Kinetic Gravity Braiding,''}
  JCAP {\bf 1010}, 026 (2010)
  doi:10.1088/1475-7516/2010/10/026
  [arXiv:1008.0048 [hep-th]].
  
\bibitem{Easson:2011zy} 
  D.~A.~Easson, I.~Sawicki and A.~Vikman,
 \textit{``G-Bounce,''}
  JCAP {\bf 1111}, 021 (2011)
  doi:10.1088/1475-7516/2011/11/021
  [arXiv:1109.1047 [hep-th]].
  
  
\bibitem{Ghersi:2016gee} 
  J.~T.~G.~Ghersi and A.~V.~Frolov,
  \textit{``Two-point correlators revisited: Fast and slow scales in multifield models of inflation,''}
  JCAP {\bf 1705}, no. 05, 047 (2017)
  doi:10.1088/1475-7516/2017/05/047
  [arXiv:1609.04770 [astro-ph.CO]].
  
  \bibitem{Barnaby:2010ke} 
  N.~Barnaby,
  \textit{``On Features and Nongaussianity from Inflationary Particle Production,''}
  Phys.\ Rev.\ D {\bf 82}, 106009 (2010)
  doi:10.1103/PhysRevD.82.106009
  [arXiv:1006.4615 [astro-ph.CO]].

\end{thebibliography}


\end{document}
